---
format: gfm
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# heritable

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://img.shields.io/cran/v/heritable.svg)](https://CRAN.R-project.org/package=heritable)
[![Codecov test coverage](https://codecov.io/gh/anu-aagi/heritable/graph/badge.svg)](https://app.codecov.io/gh/anu-aagi/heritable)
<!-- badges: end -->

`heritable` provides flexible and interpretable heritability estimation from linear mixed models in R. Given genotypes and their observed traits, it helps breeders assess whether a trial contains enough genetic signal for selection.

The package brings commonly used estimators into a unified interface and supports:

- broad-sense (`H2()`) and narrow-sense (`h2()`) heritability;
- Cullis, Oakey, Piepho, Standard, and Delta estimators, including BLUP- and BLUE-based Delta formulations;
- models fitted with [`lme4`](https://cran.r-project.org/package=lme4) and [`asreml`](https://vsni.co.uk/software/asreml-r/);
- marginal (environment-averaged) and stratified (environment-specific) heritability for genotype-by-environment models; and
- confidence intervals based on parametric bootstrap.

## Installation


The released version is available from CRAN:

``` r
install.packages("heritable")
```

The package is under active development. You can install the development version from [GitHub](https://github.com/anu-aagi/heritable) with:

``` r
# install.packages("pak")
pak::pak("anu-aagi/heritable")
```


## A simple demo

This basic example calculates broad-sense heritability for a single-environment trial. The same `H2()` interface works with `lme4` and `asreml` model objects.

```{r example}
#| output: false
#| message: false
library(heritable)

lettuce_subset <- subset(lettuce_phenotypes, loc == "L2")
fit <- lme4::lmer(y ~ rep + (1 | gen), data = lettuce_subset)
```

Uppercase `H2()` calculates broad-sense heritability. Use `target` to specify the variable encoding genetic effects,  `method` to select one or more estimators:

```{r}
#| message: false

H2(fit, target = "gen", method = c("Cullis", "Standard"))
```

`heritable` also supports multi-environment trial (MET) models. In the example below, we use `asreml` to fit a genotype-by-environment (G×E) model in which genotype, environment, and G×E interaction effects are all treated as random.

```{r example gxe}
#| output: false
#| message: false

fit_gxe <- asreml::asreml(
  fixed = y ~ rep,
  random = ~ gen + loc + gen:loc,
  trace = FALSE,
  data = lettuce_phenotypes
)
```

For G×E models, `H2()` supports multiple definitions of heritability.

The classical definition treats the G×E interaction as a nuisance source of variation. To calculate this form of heritability, set `marginal = FALSE`:

```{r}
#| message: false

H2(fit_gxe, target = "gen", marginal = FALSE)
```

Alternatively, setting `marginal = TRUE` treats the G×E interaction as part of the genetic variation expressed across environments. In this case, `H2()` estimates heritability averaged over the target environments.

```{r}
#| message: false

H2(fit_gxe, target = "gen", marginal = TRUE)
```

You can also estimate stratified heritability for a specific environment by supplying a one-row data frame to the `stratification` argument:

```{r}
#| message: false

H2(fit_gxe, target = "gen", stratification = data.frame(loc = "L1"))
```

In this example, heritability is defined and estimated specifically for environment `"L1"`.

Lowercase `h2()` estimates **narrow-sense heritability** for models that include a known additive genetic covariance structure, such as a genomic relationship matrix (GRM). In `asreml`, this is specified using `vm()`. Below, we fit a G×E model in which the additive genetic effects are modelled using the GRM.

```{r narrow-sense}
#| output: false
#| message: false

fit_grm <- asreml::asreml(
  fixed = y ~ rep,
  random = ~ vm(gen, source = lettuce_GRM) * idv(loc),
  data = lettuce_phenotypes,
  trace = FALSE
)
```

The interface of `h2()` is nearly identical to that of `H2()`. The only additional argument is `source`, which specifies the genetic covariance structure used when fitting the model. This information is required because the fitted model does not retain the covariance matrix itself.

```{r}
#| message: false

h2(fit_grm, target = "gen", source = list(lettuce_GRM = lettuce_GRM))
```

Like `H2()`, `h2()` supports both `marginal` and `stratification` arguments for G×E models. 

The returned estimate retains the fitted model, allowing parametric bootstrap confidence intervals to be obtained using the standard `confint()` generic.

```{r confidence-interval}
#| message: false

estimate <- H2(fit, target = "gen", method = c("Cullis", "Standard"))
confint(estimate, B = 100, seed = 2026)
```

See the [Get started guide](https://anu-aagi.github.io/heritable/articles/heritable.html) for further examples and the [presentation](https://anu-aagi.github.io/heritable/slides/heritable-slide-for-aagiverse-meeting.html) for the motivation and statistical framework.

## Too slow to run?

For large or complex models, heritability estimation can be computationally intensive. Runtime can be reduced by approximating the required matrix pseudoinverses and computing error-variance matrices from the mixed-model equations:

```{r}
options(exact_psd_inv = FALSE)
H2(fit, target = "gen", solver = "LMM")
```

This approach can substantially reduce runtime, although it may be less numerically stable than the default calculation. The speed improvement is generally more apparent for models with many observations, random effects, or environments.


## Support our work!

```{r}
#| warning: false
citation("heritable")
```


## Found a bug? Want to contribute? 

We welcome feedback and contributions of all shapes and sizes! Take a look at our existing [backlog](https://github.com/anu-aagi/heritable/issues), add to a relevant discussion, or lodge a new issue.
