---
format: gfm
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# heritable

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://img.shields.io/cran/v/heritable.svg)](https://CRAN.R-project.org/package=heritable)
[![Codecov test coverage](https://codecov.io/gh/anu-aagi/heritable/graph/badge.svg)](https://app.codecov.io/gh/anu-aagi/heritable)
<!-- badges: end -->

`heritable` provides flexible and interpretable heritability estimation from linear mixed models in R. Given genotypes and their observed traits, it helps breeders assess whether a trial contains enough genetic signal for selection.

The package brings commonly used estimators into a unified interface and supports:

- broad-sense (`H2()`) and narrow-sense (`h2()`) heritability;
- Cullis, Oakey, Piepho, Standard, and Delta estimators, including BLUP- and BLUE-based Delta formulations;
- models fitted with [`lme4`](https://cran.r-project.org/package=lme4) and [`asreml`](https://vsni.co.uk/software/asreml-r/);
- marginal (environment-averaged) and stratified (environment-specific) heritability for genotype-by-environment models; and
- confidence intervals based on parametric bootstrap.

## Installation


The released version is available from CRAN:

``` r
install.packages("heritable")
```

The package is under active development. You can install the development version from [GitHub](https://github.com/anu-aagi/heritable) with:

``` r
# install.packages("pak")
pak::pak("anu-aagi/heritable")
```


## A simple demo

This basic example calculates broad-sense heritability for a single-environment trial. The same `H2()` interface works with `lme4` and `asreml` model objects.

```{r example}
#| output: false
#| message: false
library(heritable)

lettuce_subset <- subset(lettuce_phenotypes, loc == "L2")
fit_lme4 <- lme4::lmer(y ~ rep + (1 | gen), data = lettuce_subset)
```

Uppercase `H2()` calculates broad-sense heritability. Use `method` to select one or more estimators:

```{r}
H2(
  fit_lme4,
  target = "gen",
  method = c("Cullis", "Oakey", "Piepho", "Delta", "Standard")
)
```

Lowercase `h2()` calculates narrow-sense heritability when the fitted model includes a known additive genetic covariance structure, such as a genomic relationship matrix. With `asreml`, this can be specified using `vm()`:

```{r narrow-sense}
#| eval: false
fit_genomic <- asreml::asreml(
  fixed = y ~ rep,
  random = ~ asreml::vm(gen, source = lettuce_GRM) * asreml::idv(loc),
  data = lettuce_phenotypes,
  trace = FALSE
)

h2(
  fit_genomic,
  target = "gen",
  marginal = TRUE,
  source = list(lettuce_GRM = lettuce_GRM)
)
```

For genotype-by-environment models, `marginal = TRUE` estimates heritability averaged across environments. Alternatively, supply a one-row data frame to `stratification` to estimate heritability within a particular environment:

```{r genotype-by-environment}
#| eval: false
H2(fit_gxe, target = "gen", marginal = TRUE)
H2(fit_gxe, target = "gen", stratification = data.frame(loc = "L1"))
```

Estimates returned by `H2()` and `h2()` retain the fitted model, allowing parametric-bootstrap confidence intervals through the standard `confint()` generic:

```{r confidence-interval}
#| eval: false
estimate <- H2(fit_lme4, target = "gen", method = c("Cullis", "Standard"))
confint(estimate, B = 500, seed = 2026)
```

See the [Get started guide](https://anu-aagi.github.io/heritable/articles/heritable.html) for further examples and the [presentation](https://anu-aagi.github.io/heritable/slides/heritable-slide-for-aagiverse-meeting.html) for the motivation and statistical framework.

## Support our work!

```{r}
#| warning: false
citation("heritable")
```


## Found a bug? Want to contribute? 

We welcome feedback and contributions of all shapes and sizes! Take a look at our existing [backlog](https://github.com/anu-aagi/heritable/issues), add to a relevant discussion, or lodge a new issue.
