---
format: gfm
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# heritable

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/heritable)](https://CRAN.R-project.org/package=heritable)
[![Codecov test coverage](https://codecov.io/gh/anu-aagi/heritable/graph/badge.svg)](https://app.codecov.io/gh/anu-aagi/heritable)
<!-- badges: end -->

`heritable` is to be the one-stop shop for heritability calculations in R. Our goal is to implement existing methods for heritability to aid reproducibility and reporting of it's calculations.`heritable` works with model outputs from `asreml` and `lme4` and support the calculations of broad and narrow sense heritability of a variety of method for single environment breeding trials. 

## Installation


Note that this package is under active development. You can install the development version of heritable from [GitHub](https://github.com/) with: 

``` r
# install.packages("pak")
pak::pak("anu-aagi/heritable")
```


## A simple demo

This is a basic example which shows you how to calculate broad-sense heritability for a single environment trial using `asreml` and `lme4`. 

```{r example}
#| output: false
#| message: false
library(heritable)  

fit_asreml <- asreml::asreml(yield ~ rep,
    random = ~ gen + rep:block,
    data = agridat::john.alpha,
    trace = FALSE
  )

fit_lme4 <- lme4::lmer(yield ~ rep + (1|gen) + (1|rep:block),
                       data = agridat::john.alpha)
```

The `H2()` function refers to broad-sense heritability and by default, it will compute all the available heritability methods for you.

```{r}
H2(fit_asreml, target = "gen")
H2(fit_lme4, target = "gen")
```

Alteratively, with the help of tidyverse functions, you can return all the measures for different models as a tibble.

```{r}
#| message: false
#| warning: false
library(tidyverse)
tibble(model = list(fit_lme4, fit_asreml)) |> 
  mutate(H2 = map(model, ~H2(.x, target = "gen"))) |> 
  unnest_wider(H2)
```

## Support our work!

```{r}
#| warning: false
citation("heritable")
```


## Found a bug? Want to contribute? 

We welcome feedback and contributions of all shapes and sizes! Take a look at our existing [backlog](https://github.com/anu-aagi/heritable/issues) and add to a relevant discussion or lodge a new issue. 
